---
title: "esa_analysis"
author: "Bryan Milstead"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

options(stringsAsFactors = FALSE) 
library(knitr)
library(vegan)
library(ecodist)
library(tidyverse)
```

## Introduction

* modified from Betty_NMSCode_20160824.R

* read taxonomy data
* read phyto counts
* join taxonomy and counts
* filter
    - !is.na(phylum)-this eliminates data without a valid genus name
    - VISIT_NO==1
    - LAKE_SAMP=='Target_Sampled'
* sum abundance by site_id and genus
* convert to wide format for NMS

```{r data, include=FALSE, echo=FALSE, cache=FALSE} 
#get the taxonomy dat
tax<-read.csv(url('https://raw.githubusercontent.com/willbmisled/lakes_database/master/output/nla_phyto_taxonomy.csv'))

#get the count data
counts<-read.csv(url('https://www.epa.gov/sites/production/files/2014-10/nla2007_phytoplankton_softalgaecount_20091023.csv'))

#merge tax and counts by genus and filter
t2007<-left_join(counts,tax,by=c("GENUS"="genus"))%>%filter(!is.na(phylum),VISIT_NO==1,LAKE_SAMP=='Target_Sampled')
    nrow(t2007) #26382
    
t2007 <- group_by(t2007,SITE_ID,GENUS)%>%
  summarise(abund=sum(ABUND, na.rm = TRUE))
    names(t2007)<-tolower(names(t2007))
        nrow(t2007) #24151
t2007<-ungroup(t2007)

phyto<-spread(t2007,genus,abund)

phyto[is.na(phyto)] <- 0 
```

* decide how many dimensions to use for the NMS

```{r data, include=FALSE, echo=FALSE, cache=FALSE} 
#Running NMS fit
phytoNMS.2<-metaMDS(phyto[,-1],k=2,trymax=500)
phytoNMS.3<-metaMDS(phytoWide[,-1],k=3,trymax=500)
phytoNMS.4<-metaMDS(phytoWide[,-1],k=4,trymax=500)

phytoNMS.5<-metaMDS(phytoWide[,-1],k=5,trymax=500)
phytoNMS.6<-metaMDS(phytoWide[,-1],k=6,trymax=500)

#save(phytoNMS.4,file="phytoNMS4.rda")
#load(file="phytoNMS4.rda")
```








load("phyto.rda")
data1<-read.csv(file="L:/Public/Betty/tox_data/Community_20160823/community_20160823/ltm_tox_20160810.csv")


a<-OTU[,-3]



```