---
title: "nlaPhytoData2007"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options("stringsAsFactors" = FALSE) 
library(RODBC)
library(dplyr)
library(tidyr)
```

## Introduction

#NOTE: This script must be run in 32bit R#

* Download the NLA 2007 phytoplankton data for ordination
* From this we will create three datasets
    - allPhytoLong:  the data as downloaded from waterbodydatabase.mdb in long format; Data definitions: SITE_ID=the NLA siteID; VISIT_NO=Ignore this, we only include the first visit; LAKE_SAMP=Ignore this also, this shows we only included data for lakes sampled; DIVISION=taxonomic division for the OTU; OTU=lowest taxonomic definition for the phytoplankter.  In most cases this it the genus; Operational Taxonomic Unit; SumOfAbund=cells per ml for the OTU.  This is the sum because sometimes the OTU comprises 2 or more lower taxonomic levels.
    - allPhytoWide=All phytoplankton data in wide format; the first column has the SITE_ID for the 2007 NLA lake; the next 260 columns are the individual abundances (cells/ml) of the individual phytoplankton OTUs
    - cyanoWide=just the cyanophyta data in wide format.  Similar to allPhytoWide but only includes the cyano OTUs.
* data saved as output/phytoDataNLA2007.rda


```{r data, include=FALSE}

#get the data
con <- odbcConnectAccess("L:/Public/Milstead_Lakes/WaterbodyDatabase/WaterbodyDatabase.mdb")

All <- sqlQuery(con, "
SELECT tblNLA_nla_phyt_softcount_20091023.SITE_ID, tblNLA_nla_phyt_softcount_20091023.DIVISION, tblNLA_nla_phyt_softcount_20091023.OTU, Sum(tblNLA_nla_phyt_softcount_20091023.ABUND) AS SumOfABUND
FROM tblNLA_nla_phyt_softcount_20091023
WHERE (((tblNLA_nla_phyt_softcount_20091023.VISIT_NO)=1) AND ((tblNLA_nla_phyt_softcount_20091023.LAKE_SAMP)='Target_Sampled'))
GROUP BY tblNLA_nla_phyt_softcount_20091023.SITE_ID, tblNLA_nla_phyt_softcount_20091023.DIVISION, tblNLA_nla_phyt_softcount_20091023.OTU;
")
close(con)
str(All)

#get the unique siteIDs for visit_no=1 and lake_samp="lake_sampled"
con <- odbcConnectAccess("L:/Public/Milstead_Lakes/WaterbodyDatabase/WaterbodyDatabase.mdb")
ids <- sqlQuery(con, "
SELECT NLA2007Sites_DesignInfo.SITE_ID
FROM NLA2007Sites_DesignInfo
WHERE (((NLA2007Sites_DesignInfo.VISIT_NO)=1) AND ((NLA2007Sites_DesignInfo.LAKE_SAMP)='Target_Sampled'))
GROUP BY NLA2007Sites_DesignInfo.SITE_ID
ORDER BY NLA2007Sites_DesignInfo.SITE_ID;
")
close(con)
str(ids)

#Delete the OTUs that are placeholders
#OTUs to delete
delOTU<-c('Chamaesiphon','Chlorophyta','Chlorophyta coccoid','Chlorophyta Filament','Chrysophyta','Cuspidothrix issatschenkoi','Cyanophyta','Diatoma','Diatoms','Diatoms (dead)','Dinoflagellates','Eucocconeis','Euglenophyte','Heterocontophyta','Psammothidium','Stigeoclonium','Unknown alga','Unknown alga flagellated')

#create allPhytoLong
allPhytoLong<-filter(All,!(OTU%in%delOTU))

#make sure all ID's included in allPhytoLong
allPhytoLong<-left_join(ids,allPhytoLong)

#create allPhytoWide
allPhytoWide <- spread(allPhytoLong[,-2], OTU, SumOfABUND)
  #replace NA with zero
    allPhytoWide[is.na(allPhytoWide)]<-0
  #remove OTU=NA
    allPhytoWide<-allPhytoWide[,-262]
    
dim(allPhytoWide) #1152 262
sum(allPhytoWide[,-1],na.rm=TRUE) #42311685
    

#create cyanoWide
  cyanoWide<-spread(filter(allPhytoLong, DIVISION=="Cyanophyta")[,-2], OTU, SumOfABUND)
  #make sure all ID's included 
    cyanoWide<-left_join(ids,cyanoWide)
  #replace NA with zero
    cyanoWide[is.na(cyanoWide)]<-0
  #remove OTU=NA
    cyanoWide<-cyanoWide[,-262]
    
#save data as phytoDataNLA2007.rda
    save(allPhytoLong,allPhytoWide,cyanoWide, file='phytoDataNLA2007.rda')
    
```

## Data Steps

* get the phytoplankton data
    - restrict to VISIT_NO==1 and LAKE_SAMP=='Target_Sampled'
    - ABUND-summed by OTU to combine multiple taxa within an OTU
* The following OTU's will be deleted.  These are unidentified taxa. 

```{r delOTU, echo=FALSE}
delOTU
```

* create allPhytoWide-a dataset with the SITE_ID, and the OTUs as the column names.  Values are the abundances for each OTU
    - replace NA values with zeros
* create cyanoWide-similar to allPhytoWide but with just the cyanophyta OTUs.
    - replace NA values with zeros
* rename All to allPhytoLong
* save allPhytoLong, allPnytoWide, and cyanoWide as output/phytoDataNLA2007.rda

##vegan
* compare dissimilarity indices

```{r vegan, echo=FALSE}
#create community matrix
  c1<-cyanoWide[,-1]
  row.names(c1)<-cyanoWide[,1]
  
  p1<-allPhytoWide[,-1]
  row.names(p1)<-allPhytoWide[,1]

t1=metaMDS(p1, k=3,tryMax=20) 
plot(t1)
ordiplot(t1,type="p",choices=c(1,3))
orditorp(t1,display="species",col="red",air=0.01)
orditorp(t1,display="sites",cex=1.25,air=0.01)

par(mfrow=c(2,2))
ordiplot(t1,type="p",choices=c(1,2))
ordiplot(t1,type="p",choices=c(1,3))
ordiplot(t1,type="p",choices=c(2,3))

t1
str(t1)

summary(t1$points)

 
```
