---
title: "phytoData"
author: "B"
date: "Tuesday, October 27, 2015"
output: html_document
---
```{r setup, include=FALSE, echo=FALSE, cache=FALSE} 
#########function to install (if needed) and load R packages by list
libs<-c("RODBC","dplyr") #list of packages to load

installLoad<-function(pck){ #user defined function
    if(!pck%in%installed.packages()){install.packages(pck)}
    require(pck, character.only = TRUE)
  }
lapply(libs,function(x) installLoad(x))  #Load/Install require packages
```

To Do List
-------------------------
* check INDXSAMP_PHYT logic (see below)
* add environmental data


Introduction
-------------------------
Work with the NLA 2007 phytoplankton data

Data Steps
-------------------------

* get the phytoplankton data from waterbodydatabase.mdb

```{r phytoData, include=FALSE, echo=FALSE, cache=FALSE} 
# Read data-****Make Sure the Path Is Correct****
con <- odbcConnectAccess("c:/bryan/EPA/Data/WaterbodyDatabase/WaterbodyDatabase.mdb")
phyto<- sqlQuery(con, "
SELECT tblNLA_nla_phyt_softcount_20091023.SITE_ID, tblNLA_nla_phyt_softcount_20091023.VISIT_NO, tblNLA_nla_phyt_softcount_20091023.DATE_COL, tblNLA_nla_phyt_softcount_20091023.SAMPLED_PHYT, tblNLA_nla_phyt_softcount_20091023.SAMPLE_CATEGORY, tblNLA_nla_phyt_softcount_20091023.INDXSAMP_PHYT, tblNLA_nla_phyt_softcount_20091023.SAMPLE_DEPTH, tblNLA_nla_phyt_softcount_20091023.TAXATYPE, tblNLA_nla_phyt_softcount_20091023.OTU, tblNLA_nla_phyt_softcount_20091023.ABUND, tblNLA_nla_phyt_softcount_20091023.DIVISION, tblNLA_nla_phyt_softcount_20091023.FAMILY, tblNLA_nla_phyt_softcount_20091023.ORDER, tblNLA_nla_phyt_softcount_20091023.GENUS, tblNLA_nla_phyt_softcount_20091023.SPECIES, tblNLA_nla_phyt_softcount_20091023.VARIETY
FROM tblNLA_nla_phyt_softcount_20091023;
")
close(con)
#str(phyto)
```

* Data Filters

    - **INDXSAMP_PHYT=='YES'**: NO=3776, YES=28195; This appears to be a field to identify which of multiple samples to use.  Additional samples collected as field duplicates (SAMPLE_CATEGORY=='D'), and potentially as VISIT_NO==2. 
    - **SAMPLED_PHYT="YES"**; 3 lost samples removed
* Check to see there is only one VISIT_NO per SITE_ID; there is.
* Check to see there is only one ABUND per OTU & SITE_ID; 
    - There are 1606 SITE_ID OTU combinations with multiple ABUND measures
    - This happens when there are more than one taxa (e.g. two species in an OTU defined as a genus)
    - Fix this by summing ABUND by SITE_ID and OTU
    - create new data.frame "OTU" with the aggregated data
* save data as an r dataset

```{r filters, include=FALSE, echo=FALSE, cache=FALSE} 
#explore the data first
  table(phyto$INDXSAMP_PHYT) #NO=3776, YES=28195
  length(unique(phyto[,'SITE_ID']))  #1157
  length(unique(phyto[phyto$INDXSAMP_PHYT=='YES','SITE_ID'])) #1157
  
  table(phyto[phyto$INDXSAMP_PHYT=='YES','SAMPLE_CATEGORY']) #all are "P"
  table(phyto[phyto$INDXSAMP_PHYT=='YES','SAMPLED_PHYT']) #3 were lost
  table(phyto[phyto$INDXSAMP_PHYT=='YES','VISIT_NO']) #visit==1 28195
#filter
  #INDXSAMP_PHYT=='YES'
    nrow(phyto) #31971
    phyto<-phyto[phyto$INDXSAMP_PHYT=='YES',]
    nrow(phyto) #28195
  #SAMPLED_PHYT="YES"
    phyto<-phyto[phyto$SAMPLED_PHYT=="YES",]
    nrow(phyto) #27192

#check that there is only one VISIT_NO per site
q<-unique(phyto[,c('SITE_ID','VISIT_NO')])
b<-tally(group_by(q,SITE_ID, VISIT_NO))
nrow(b[b$n>1,]) #0

#check that there is only one "ABUND" per SITE_ID and OTU
b<-tally(group_by(phyto,SITE_ID, OTU))
nrow(b)
nrow(b[b$n>1,]) #1606 observations with more than one value for siteID

#for SITE_ID/OTU combinations with multiple observations sum the results
d<-summarise(group_by(phyto,SITE_ID, OTU), sumAbund=sum(ABUND))

#create new data.frame "OTU" with the aggregated data
OTU<-merge(b,d,by=c('SITE_ID','OTU'))

#save
  save(phyto,OTU,file='data/phyto.rda')
```



Data Definitions
-------------------------

* to access data use: load('data/phyto.rda')

**data.frame: phyto obs.=`r nrow(phyto)`**

Field  | Units | Description
------------- | ------------- | -------------
**SITE_ID** | (char) |Unique site identification number
**VISIT_NO** | (num) |Lake visit number. 1-1st field sampling visit; 2-2nd field sampling visit; after filtering only sample 1 will remain.
**DATE_COL** | (date) |Date of site visit
**SAMPLED_PHYT** | (char) |Phytoplankton sample collected?
**SAMPLE_CATEGORY** | (char) |(P)rimary or Field (D)uplicate
**INDXSAMP_PHYT** | (YES/NO) |Index phytoplankton sample for site (YES/NO)-typically the Visit 1 Primary sample
**SAMPLE_DEPTH** | (m) |Depth of integrated sample (typically 2 m)
**TAXATYPE** | (cells/ml) |Type of algae
**OTU** | (char) |Operational Taxonomic Unit 
**ABUND** | (cells/mL) |Abundance
**DIVISION** | (char) |Division name (from taxa list)
**ORDER** | (char) |Order name (from taxa list)
**FAMILY** | (char) |Family name (from taxa list)
**GENUS** | (char) |Genus name (from taxa list)
**SPECIES** | (char) |Species name (from taxa list)
**VARIETY** | (char) |Variety name (from taxa list)

**data.frame: OTU obs.=`r nrow(OTU)`**

Field  | Units | Description
------------- | ------------- | -------------
**SITE_ID** | (char) |Unique site identification number
**OTU** | (char) |Operational Taxonomic Unit
**n** | (num) | number of observations in sumAbund
**sumAbund** | (cells/mL) |Abundance of OTU; some OTU include multiple observations of lower taxonomic levels.  These are summed.


  


    
